### 2014-04-15 thesis-proposal/acc.tex

% !TEX root = omar-thesis-proposal.tex
\section{Active Code Completion}\label{acc}
A language's syntax and semantics influences the behavior of a variety of tools beyond the compiler. For example, software developers today make heavy use of the code completion support found in modern source code editors  \cite{murphy_how_2006}. Editors for object-oriented languages provide code completion in the form of a floating menu containing  contextually-relevant variables, fields, methods, types and other code snippets. By navigating and selecting from this menu, developers are able to avoid many common spelling and logic errors, eliminate unnecessary keystrokes and explore unfamiliar APIs without incurring the mental overhead associated with switching to an external documentation tool.
% The way code completion menus brought documentation systems into the editor, active code completion interfaces bring external tools into the editor.

@@

\begin{lstlisting}
  public Color getDefaultColor() {
      return _
\end{lstlisting}

\begin{figure*}\label{color}
\begin{center}
\includegraphics[width=40pc]{color_palette.png}\end{center}
\caption{(a) An example code completion palette associated with the \texttt{Color} class. (b) The source code generated by this palette.}
\label{colorpalette}
\end{figure*}

If the developer invokes the code completion command at the indicated cursor position (\li{_}), the editor  looks  for a {\it palette definition} associated with the {\it type} of the expression being entered, which in this case is  \verb|Color|. If an associated palette is found, a menu item briefly describing this palette is added to the standard code completion menu. When selected, the corresponding palette is shown, replacing the standard code completion menu. Figure \ref{colorpalette}(a) gives an example of a simple palette that may be associated with the \verb|Color| class\footnote{A video demonstrating this process is available at \url{http://www.cs.cmu.edu/~NatProg/graphite.html}.}.

The developer can interact with such palettes to provide parameters and other information related to their intent, and receive immediate feedback about the effect these choices will have on the behavior of the object being constructed. When this interaction is complete, the palette generates appropriate source code for insertion at the cursor. Figure \ref{colorpalette}(b) shows the inserted code after the user presses \textsc{enter}.

We sought to address the following questions before designing and implementing our active code completion system:

@@

\subsection{Remaining Tasks and Timeline}
This work has been published at ICSE 2012 \cite{omar-graphite12}, and we do not plan on extending it further in this thesis. The main tasks have to do with relating it to the mechanisms in the previous sections, so that it generalizes beyond Java. We plan to do this when writing the thesis. There are also some pieces of related work that have been published since this paper was accepted that we need to review.

>>>>

% !TEX root = omar-thesis-proposal.tex
\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont, 
commentstyle=\itshape\ttfamily, 
stringstyle=\ttfamily,
numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML,moredelim=[il][\sffamily]{?},mathescape=true,showspaces=false,showstringspaces=false,xleftmargin=15pt, morekeywords=[1]{tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,toast},deletekeywords={for,double,in},classoffset=0,belowskip=\smallskipamount,
moredelim=**[is][\color{cyan}]{SSTR}{ESTR},
moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
moredelim=**[is][\color{purple}]{SCSS}{ECSS},
moredelim=**[is][\color{brown}]{SSQL}{ESQL},
moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
moredelim=**[is][\color{gray}]{SNAT}{ENDNAT}, 
moredelim=**[is][\color{blue}]{SURL}{EURL},
moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
moredelim=**[is][\color{YellowGreen}]{SID}{EID},
moredelim=**[is][\color{Sepia}]{SUS}{EUS}
}
\section{Active Code Completion}\label{acc}
An abstraction might benefit from support from a variety of tools beyond the compiler. For example, software developers today make heavy use of the code completion support found in modern source code editors  \cite{murphy_how_2006}. Editors for object-oriented languages provide code completion in the form of a floating menu containing  contextually-relevant variables, assignables, fields, methods, types and other code snippets. By navigating and selecting from this menu, developers are able to avoid many common spelling and logic errors, eliminate unnecessary keystrokes and explore unfamiliar APIs without incurring the mental overhead associated with switching to an external documentation tool.
% The way code completion menus brought documentation systems into the editor, active code completion interfaces bring external tools into the editor.

@@

\begin{lstlisting}[language=Java]
  public Color getDefaultColor() {
      return _
\end{lstlisting}

\begin{figure*}\label{color}
\begin{center}
\includegraphics[width=40pc]{color_palette.png}\end{center}
\caption{(a) An example code completion palette associated with the \texttt{Color} class. (b) The elaboration generated by this palette.}
\label{colorpalette}
\end{figure*}

If the developer invokes the code completion command at the indicated cursor position (\li{_}), the editor  looks  for a {\it palette definition} associated with the {\it type} that the expression being entered is being analyzed against, which in this case is  \verb|Color| (due to the return type annotation on the method). If an associated palette is found, a menu item briefly describing this palette is added to the standard code completion menu. When selected, the corresponding palette is shown, replacing the standard code completion menu. Figure \ref{colorpalette}(a) gives an example of a simple palette that may be associated with the \verb|Color| class\footnote{A video demonstrating this process is available at \url{http://www.cs.cmu.edu/~NatProg/graphite.html}.}.

The developer can interact with such palettes to provide parameters and other information related to their intent, and receive immediate feedback about the effect these choices will have on the behavior of the object being constructed. When this interaction is complete, the palette produces an elaboration for insertion at the cursor, of the type (here, class) that the palette was associated with (Figure \ref{colorpalette}(b)).

This is, in essence, an enriched edit-time variant of the mechanism used in Wyvern. As discussed in Sec. \ref{aparsing}, we associate a palette by providing metadata in the form of a class annotation. For example, we might write:
\begin{lstlisting}[language=Java]
@GraphitePalette(url="http://cs.cmu.edu/~comar/color-palette.html")
class Color { ... }
\end{lstlisting}

Were a type-aware editor for Wyvern written, it might instead use metadata:
\begin{lstlisting}
objtype Color
  ...
  metadata : HasPalette = new
    val palette_url = <SURLhttp://cs.cmu.edu/~comar/color-palette.htmlEURL>
\end{lstlisting}

\subsection{Design Process}
We sought to address the following questions before designing and implementing our active code completion system:

@@

\subsection{Remaining Tasks and Timeline}
This work has been published at ICSE 2012 \cite{Omar:2012:ACC:2337223.2337324}. The main remaining tasks have to do with specifying it in terms of the bidirectional typing mechanisms in Sec. 3 more directly, rather than informally as in our paper. We plan to do this when writing the dissertation. There are also some pieces of related work that have been published since this paper was accepted that we need to review.

### 2014-04-15 thesis-proposal/background.tex

More powerful and direct compile-time \emph{term rewriting mechanisms} available in some languages can also be used for optimization, as well as for introducing specialized error checking logic and implementing new abstractions. For example, typed macros, such as those in MetaML \cite{Sheard:1999:UMS}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala \cite{scala-macros}, take full control over all of the code that they enclose. This can be problematic, however, as outer macros can interfere with inner macros. Moreover, once a value escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope, because the output of a macro is simply a term in the underlying language (a problem fundamentally related to the problem of relying on a common intermediate language,  described in Section \ref{external-approaches}). Thus, macros represent at best a partial solution: they can be used to automate code generation, but not to globally extend the syntax or static guarantees of a language. It can also be difficult to reason about the semantics of code when any number of enclosing macros may be manipulating it, and to build tools that operate robustly in their presence.

>>>>

More powerful and direct compile-time \emph{term rewriting mechanisms} available in some languages can also be used for optimization, as well as for introducing specialized error checking logic and implementing new abstractions. For example, typed macros, such as those in MetaML \cite{Sheard:1999:UMS}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala \cite{ScalaMacros2013}, take full control over all of the code that they enclose. This can be problematic, however, as outer macros can interfere with inner macros. Moreover, once a value escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope, because the output of a macro is simply a term in the underlying language (a problem fundamentally related to the problem of relying on a common intermediate language,  described in Section \ref{external-approaches}). Thus, macros represent at best a partial solution: they can be used to automate code generation, but not to globally extend the syntax or static guarantees of a language. It can also be difficult to reason about the semantics of code when any number of enclosing macros may be manipulating it, and to build tools that operate robustly in their presence.
